---
title: "Simulations Done Right"
author: "Alejandro Schuler"
execute: 
    echo: true
    eval: true
format:
  revealjs:
    theme: default
    slide-number: true
    code-line-numbers: false
    code-overflow: wrap
    highlight-style: github
    smaller: true
    scrollable: true
---

## How to do Simulations

```{dot}
//| echo: false
digraph {
  rankdir=TB
  node [shape=box, style="rounded,filled", fontname="Helvetica", fontsize=18, margin="0.3,0.2"]
  
  A [label="1. Paper Claims 🎯", fillcolor="#f88674ff"]
  B [label="2. Simulation Goals 🎯", fillcolor="#f88674ff"]
  C [label="3. Mockup Outline, Tables & Figures 📊", fillcolor="#f6cf71"]
  D [label="4. DGPs & Settings ⚙️", fillcolor="#66c5cc"]
  E [label="5. Implement ⚙️", fillcolor="#66c5cc"]
  
  A -> B -> C -> D -> E [penwidth=2, color=grey40]
  E -> A [penwidth=1.5, color=grey60, style=dashed, constraint=false]
}
```

## Top Tips

:::: {.columns}

::: {.column width="30%" style="background: #f88674ff"}
#### 🎯 Thoughtful Goals = Direction!

- practical utility
- key insight
- audience
:::

::: {.column width="30%" style="background: #f6cf71"}
#### 📊 Communication = Focus!

- structure + organize
- minimize cognitive load
:::

::: {.column width="30%" style="background: #66c5cc"}
#### ⚙️ Clean, Fast Sims = Productivity!

- simple DGPs
- minimize compute
- modular code
:::

::::

# 🎯 Setting Goals {background-color="#f88674ff"}

## Setting Goals

::: {.incremental}
- paper = claims 🎯 + evidence 
    - evidence = simulations 📊, applications 🗂️, theory ♾️, etc.
- A good claim likely...
    - has practical utility
    - builds insight
    - is honest and well-scoped
- Many claims are possible. Prioritize:
    - relevance to audience
    - potential to change practice or thinking
- sims often support claims like: *"[method] works [well] when [condition] and not when [other condition]"*
::: 

## Exercise: TMLE

> *You wake up. It's 2006. You look in the mirror and you see you are Mark van der Laan. You've just invented TMLE.*

- What are the claims you should set out for the TMLE paper? Which of them can be supported by theory and which by simulations?

## Some Claims

- **(theory)** TMLE is a general method for constructing efficient estimators of pathwise differentiable estimands
    - Estimating equations already does this, so while this needs to be shown, it's not enough to prove TMLE is useful!

- TMLE is in practice "better" than...
    - Plug-in estimators? AIPW is already better
    - AIPW?

. . .

❓ What is meant by "better"? When should it be better and when worse?

- It works in the real world? Computationally tractable, doesn't require specific data types, etc. (usually addressable w/ theory and a single real data application)

## Simulation Goals (Initial)

1. Across the board, AIPW and TMLE behave similarly in large samples when they have access to the same learners
2. TMLE is better than plugins, allowing for good coverage of Wald intervals
3. There are cases where TMLE is better than AIPW in small samples

## Simulation Goals (Some Questions)

1. Across the board, AIPW and TMLE behave similarly in large samples when they have access to the same learners
    - What learners?
    - By what metric(s)?
2. Like AIPW, TMLE is better than plugins, allowing for good coverage of Wald intervals
    - What learners?
    - What should $n$ be here?
    - Presumably using IF-based CI
    - Metric?
    - But is there a way to get CIs for the plugins as a naive baseline?
3. There are cases where TMLE is better than AIPW in small samples
    - Can we back these up with some kind of theory?
    - In what DGPs?

## Simulation Goals (v2)

1. Across the board, AIPW and TMLE behave similarly in terms of point estimate and SE estimate in large samples when they have access to the same learners
2. Like AIPW, TMLE improves the MSE of plugins in moderate sample sizes and allows for good coverage of Wald intervals based on IF SEs, even compared to the same SE put around the plugin
3. TMLE is better than AIPW in small samples when the outcome is bounded (?)

# 📊 Presentation {background-color="#f6cf71"}


## Mockup Evidence for Goal 1 {.smaller}

> Across the board, AIPW and TMLE behave similarly in terms of point estimate and SE estimate in large samples when they have access to the same learners

```{r}
#| echo: false
library(tidyverse)

set.seed(42)

mockup <- expand_grid(
    n = c(100, 500, 1000, 5000, 10000),
    DGP = c("Simple", "Moderate", "Complex"),
    Estimand = c("Point Estimate", "SE Estimate")
) |>
    mutate(
        complexity = match(DGP, c("Simple", "Moderate", "Complex")),
        base = complexity * 0.3 / sqrt(n),
        abs_diff = abs(base + rnorm(n(), 0, base * 0.15)),
        DGP = factor(DGP, levels = c("Simple", "Moderate", "Complex"))
    )

ggplot(mockup, aes(x = n, y = abs_diff, color = DGP)) +
    geom_line(linewidth = 1) +
    geom_point(size = 2.5) +
    scale_x_log10() +
    facet_wrap(~Estimand) +
    labs(
        title = "Goal 1: AIPW and TMLE converge as n grows",
        x = "Sample size (n)",
        y = "|AIPW - TMLE| (avg over reps)",
        color = "DGP"
    ) +
    theme_minimal()
```

## Mockup Evidence for Goal 2 {.smaller}

> Like AIPW, TMLE improves the MSE of plugins in moderate sample sizes and allows for good coverage of Wald intervals based on IF SEs.

Fix $n = 300$, do $B$ reps from the same 3 DGPs, tables like:

| | plugin | AIPW | TMLE |
|---|---|---|---|
| MSE | | | |
| bias | | | |
| variance | | | |
| coverage | | | |

One per each DGP (or 3 sub-tables). Probably just ATE, send the other estimand to the appendix.

- Learners? Some fast nonparametric thing, not too relevant to goal what it is

## Mockup Evidence for Goal 3

> TMLE is better than AIPW in small samples when the outcome is bounded (?)

The evidence for this goal might be a subset of evidence for the above. So just suggests that one of our DGPs should have a bounded outcome, and then we can point to the relevant rows in the above table to support this claim.


## Telling the Story

For each bit of evidence you present, you should **tell the reader** why it is there. Tell them your goal! Then give them the evidence. Then tell them how the evidence supports the goal. Be prosaic and direct.

. . .

> "In the first part I tell 'em what I am going to tell 'em; in the second part — well, I tell 'em; in the third part I tell 'em what I've told 'em."

::: {.callout-note}
## Example
The point of this application is to show how fast TMLE is on a real-world dataset of size *blah blah* and to prove that it works with mixed datatypes, demonstrate how it can be prespecified, and produces plausible results.
:::

## ADEMP Framework

:::: {.columns}

::: {.column width="40%"}
From [Morris et al. (2019)](https://pubmed.ncbi.nlm.nih.gov/30652356/)

*"Using simulation studies to evaluate statistical methods"*

*Statistics in Medicine*, 38(11), 2074–2102.
:::

::: {.column width="60%"}
::: {.incremental}
Before presenting results, clearly describe:

- **A**ims: what are the claims you are trying to support?
- **D**ata-generating mechanisms: how did you simulate the data? → `DGPs`
- **E**stimands: what are the true values? → e.g. `ATE()` helper
- **M**ethods: what methods are you using? → `ESTIMATORS`
- **P**erformance measures: what metrics? → the `summarize` step and your tables/figures
:::
:::

::::

. . .

* Aims is most important, others are in service of that
* This is part of the *"tell them what you're going to tell them."*
* Can do separately for each claim or once

## Example from a Recent Paper

[arxiv.org/pdf/2503.22284](https://arxiv.org/pdf/2503.22284)

# ⚙️ Implementation {background-color="#66c5cc"}

## Advice for Machine Learning

- Avoid super-learner and cross-validation
    - use single-split, 1-2 good learners
- Understand learners and hyperparams

## ML Method Recommendations {.smaller}

| Method | Speed | Notes |
|---|---|---|
| **MARS** (`earth`) | ⚡ Super fast | Works well with a single set of hyperparams. Ensure depth is high enough, enable pruning, allow interaction. |
| **Random forests** | 🏃 Fast | Works well with default hyperparams. Worse for smooth functions. |
| **GBT** (lightgbm/xgboost) | 🏃 Fast | Beats almost everything. Requires tuning over # trees, depth, learning rate. Use early stopping. |
| **Elastic net** | 🏃 Fast (ridge) | Loses to others with moderate nonlinearity. Good for baselines. Requires tuning regularization. |
| **KRR** (kernel ridge) | 🐌 Slow | Good for smooth functions, easy to analyze theoretically. Use only for $n < 1000$. |

## ML Method Recommendations 

- **Deep learning?** 
    - Avoid unless you specifically need a differentiable model, custom loss function, fine-tuning, etc.
    - Hard to tune
    - xgboost wins for tabular data.
- **HAL?** lol. lmao even

## Advice for DGPs

Being able to *reason* about your simulations and *iterate* on them is much more important than making them realistic. For example:

::: {.incremental}
- $X \sim \mathsf{Unif}([-1,1]^d)$ unless otherwise required. Bounded domain helps reason about function ranges
- $A \sim \mathsf{Bern}(\text{expit}(\rho(X)))$. Keep $\rho$ simple + sparse and, unless empirical positivity violations are your goal, shoot for $\rho \in [-3,3]$
- $Y \sim \mu(A, X) + \mathsf{N}(0,\sigma)$. Keep $\mu$ simple + sparse.
:::

. . . 

Pay attention to:

::: {.incremental}
- positivity violations and dist of $A|X$
- percent of outcome variance explainable by $\mu$ (signal-to-noise ratio)
- percent of variance of $\mu$ explainable by best linear approximation 
:::

## Evidence for Goal 1 {.smaller}

> Across the board, AIPW and TMLE behave similarly in terms of point estimate and SE estimate in large samples when they have access to the same learners

```{r}
#| echo: false
library(tidyverse)

set.seed(42)

mockup <- expand_grid(
    n = c(100, 500, 1000, 5000, 10000),
    DGP = c("Simple", "Moderate", "Complex"),
    Estimand = c("Point Estimate", "SE Estimate")
) |>
    mutate(
        complexity = match(DGP, c("Simple", "Moderate", "Complex")),
        base = complexity * 0.3 / sqrt(n),
        abs_diff = abs(base + rnorm(n(), 0, base * 0.15)),
        DGP = factor(DGP, levels = c("Simple", "Moderate", "Complex"))
    )

ggplot(mockup, aes(x = n, y = abs_diff, color = DGP)) +
    geom_line(linewidth = 1) +
    geom_point(size = 2.5) +
    scale_x_log10() +
    facet_wrap(~Estimand) +
    labs(
        title = "Goal 1: AIPW and TMLE converge as n grows",
        x = "Sample size (n)",
        y = "|AIPW - TMLE| (avg over reps)",
        color = "DGP"
    ) +
    theme_minimal()
```

## Settings for Goal 1 {.smaller}

> Across the board, AIPW and TMLE behave similarly in terms of point estimate and SE estimate in large samples when they have access to the same learners

::: {.incremental}
- What DGPs? What estimand(s)? What learners?
- Do ~100 reps first, just get the vibe (do 2 reps first to see if code even runs!)
- $n$ from 100 to 10k? Start with 3–4 increments, add more later
- Maybe 3 DGPs of increasing complexity/confounding. Simple $W,A,Y$ setup w/ binary $A$ for intelligibility. Doesn't matter much for this goal
- For convenience, choose estimands that are computationally easy and work with the same DGP, e.g. ATE and incremental pscore. Having 2 estimands helps enforce the top-level generality claim, but stick to ATE for first round of output
- Learners: fast nonparametric learners for both OR and PS
:::

## Advice for Simulation Code

::: {.incremental}
- Separate **DGPs** are separate objects with `draw` method 
- **Estimators** are functions, separate, but can share innards
    - separate out learning of common nuisances 
    - Generally return `(est, est_se)`
    - Clear claims will help debug!
- **Simulation loop** over DGPs, estimators, and reps is a simple nested loop that produces a tidy data frame of results
    - Use `furrr` for drop-in parallelization from `map` in R, parallelize across *reps*
- Keep DGPs (and helpers) in one file, estimators (and helpers) in another, and run your sims/tests in a notebook, saving results to disk as needed
:::

## DGPs: Simplest Version

```{r}
#| echo: true
library(tidyverse)
library(magrittr)

draw <- \(n) tibble(
    W1 = runif(n, -1, 1),
    W2 = runif(n, -1, 1),
    A = rbinom(n, 1, plogis(W1 + W2)),
    Y = W1 + W2 + A + rnorm(n)
)
```

## DGPs: More Modular

```{r}
#| echo: true
rho <- \(w1, w2) w1 + w2
mu <- \(a, w1, w2) w1 + w2 + a

draw <- \(n) tibble(
    W1 = runif(n, -1, 1),
    W2 = runif(n, -1, 1),
    A = rbinom(n, 1, plogis(rho(W1, W2))),
    Y = mu(A, W1, W2) + rnorm(n)
)
```

## DGPs: OOP Approach {.smaller}

```{r}
#| echo: true
simple_dgp <- structure(
    list(
        rho = \(w1, w2) w1 + w2,
        mu = \(a, w1, w2) w1 + w2 + a,
        sigma = 1
    ),
    class = "dgp"
)

complex_dgp <- structure(
    list(
        rho = \(w1, w2) w1 + w2 + w1 * abs(w2),
        mu = \(a, w1, w2) w1 + w2 + abs(w2) + 0.5 * w1 * w2,
        sigma = 1
    ),
    class = "dgp"
)

draw <- function(x, ...) UseMethod("draw")
draw.dgp <- \(dgp, n) dgp %$% tibble(
    W1 = runif(n, -1, 1),
    W2 = runif(n, -1, 1),
    A = rbinom(n, 1, plogis(rho(W1, W2))),
    Y = mu(A, W1, W2) + rnorm(n, 0, sigma)
)

draw_intervene <- function(x, ...) UseMethod("draw_intervene")
draw_intervene.dgp <- \(dgp, n, treatment) dgp %$% tibble(
    W1 = runif(n, -1, 1),
    W2 = runif(n, -1, 1),
    A = treatment(W1, W2),
    Y = mu(A, W1, W2) + rnorm(n)
)
```

## DGPs: Usage

```{r}
#| echo: true
draw(simple_dgp, 3)
draw(complex_dgp, 3)
draw_intervene(simple_dgp, n=3, treatment = \(w1, w2) 1)
```

## DGP Helpers {.smaller}

```{r}
#| echo: true
ATE <- function(x, ...) UseMethod("ATE")
ATE.dgp <- \(dgp, n = 1e6) {
    mu0 <- dgp |> draw_intervene(n, \(w1, w2) 0) %$% mean(Y)
    mu1 <- dgp |> draw_intervene(n, \(w1, w2) 1) %$% mean(Y)
    mu1 - mu0
}

R2 <- function(x, ...) UseMethod("R2")
R2.dgp <- \(dgp, n = 1e6) {
    pop <- dgp |>
        draw(n) |>
        mutate(mu = dgp$mu(A, W1, W2))
    mu_lin <- lm(mu ~ A + W1 + W2, data = pop)
    pop <- pop |>
        mutate(mu_lin = predict(mu_lin, pop))

    list(
        R2     = pop %$% { var(mu) / var(Y) },
        R2_lin = pop %$% { var(mu_lin) / var(mu) }
    )
}
```

```{r}
#| echo: true
ATE(simple_dgp)
R2(simple_dgp)
```

## Estimators: General Structure

```{r}
#| eval: false
my_estimator <- function(
    data, # always a data frame with consistently named columns
    hyperparams_or_nuisances # whatever else you need
) {
    # computations ...

    list(
        est    = ..., # point estimate
        est_se = ...  # SE estimate
    )
}
```

So that it can be used like this:

```{r}
#| eval: false

dgp |> draw(100) |> my_estimator()
```

- write code for simulation, not for a human user

## Estimator: Difference in Means

```{r}
diff_means <- function(data, ...) {
    data |>
        group_by(A) |>
        summarize(
            mu_hat  = mean(Y),
            var_hat = var(Y),
            n       = n()
        ) %$%
        list(
            est    = mu_hat[A == 1] - mu_hat[A == 0],
            est_se = sqrt(var_hat[A == 1] / n[A == 1] +
                          var_hat[A == 0] / n[A == 0])
        )
}
```



## Estimators: TMLE & AIPW {.smaller}

```{r}
library(zeallot)

tmle <- function(data, nuisances) {
    c(pi, alpha, mu) %<-% nuisances

    eps <- lm(Y - mu(A, W1, W2) ~ alpha(A, W1, W2) - 1, data) |> coef() |> unname()
    data |>
        mutate(
            mu1 = mu(1, W1, W2) + eps * alpha(1, W1, W2),
            mu0 = mu(0, W1, W2) + eps * alpha(0, W1, W2),
            phi = alpha(A, W1, W2) * (Y - mu(A, W1, W2)) + (mu1 - mu0)
        ) %$%
        list(
            est    = mean(mu1 - mu0),
            est_se = sd(phi) / sqrt(nrow(data))
        )
}
```

. . . 

```{r}
aipw <- function(data, nuisances) {
    c(pi, alpha, mu) %<-% nuisances

    data |>
        mutate(
            phi = alpha(A, W1, W2) * (Y - mu(A, W1, W2)) +
                  (mu(1, W1, W2) - mu(0, W1, W2))
        ) %$%
        list(
            est    = mean(phi),
            est_se = sd(phi) / sqrt(nrow(data))
        )
}
```

## Learning Nuisances {.smaller}

```{r}
library(tidymodels)

fit_learners <- function(data) {
    # Convert A to factor for classification
    data <- data |> mutate(A = factor(A))
    
    pi_obj <- mars(prod_degree = 3, num_terms = 50) |>
        set_engine("earth") |>
        set_mode("classification") |>
        fit(A ~ W1 + W2, data)

    mu_obj <- mars(prod_degree = 3, num_terms = 50) |>
        set_engine("earth") |>
        set_mode("regression") |>
        fit(Y ~ A + W1 + W2, data)

    pi    = \(w1, w2) predict(pi_obj, tibble(W1 = w1, W2 = w2), type = "prob")$.pred_1
    alpha = \(a, w1, w2) a / pi(w1, w2) -
                            (1 - a) / (1 - pi(w1, w2))
    mu    = \(a, w1, w2) predict(mu_obj, tibble(A = factor(a, levels = c(0, 1)), W1 = w1, W2 = w2))$.pred

    list(pi = pi, alpha = alpha, mu = mu)
}
```

## Sample Splitting

```{r}
# Train on a separate sample (approximation of cross-fitting,
# but much faster and less complicated)
nuisances <- simple_dgp |> draw(500) |> fit_learners()
data <- simple_dgp |> draw(500)
```

. . . 

```{r}
list(
    dm   = diff_means(data),
    tmle = tmle(data, nuisances),
    aipw = aipw(data, nuisances)
) |> map_dfr(as_tibble, .id="method")
```

## Simulation Loop {.smaller}

```{r}
REPS <- 20
N <- 500

DGPS <- list(simple = simple_dgp, complex = complex_dgp)
ESTIMATORS <- list(dm = diff_means, tmle = tmle, aipw = aipw)

1:REPS |> map(\(rep) {
    DGPS |> imap(\(dgp, dgp_name) {
        nuisances <- dgp |> draw(N) |> fit_learners()
        data <- dgp |> draw(N)

        ESTIMATORS |> imap(\(estimator, estimator_name) {
            res <- estimator(data, nuisances)
            tibble(
                rep       = rep,
                DGP       = dgp_name,
                estimator = estimator_name,
                est       = res$est,
                se        = res$est_se
            )
        }) |> bind_rows()
    }) |> bind_rows()
}) |> bind_rows() -> results
```

- Save the raw `results` object to file, possibly with a pointer to the commit hash for replicability
- If you optimize runtime, you don't have to worry so much about saving carefully since it's so fast to regenerate!

## Parallelize with `furrr` {.smaller}

```{r}
#| exec: false

library(furrr)
plan(multicore, workers = 10)

1:REPS |> furrr::future_map(\(rep) {
    DGPS |> imap(\(dgp, dgp_name) {
        nuisances <- dgp |> draw(N) |> fit_learners()
        data <- dgp |> draw(N)

        ESTIMATORS |> imap(\(estimator, estimator_name) {
            res <- estimator(data, nuisances)
            tibble(
                rep       = rep,
                DGP       = dgp_name,
                estimator = estimator_name,
                est       = res$est,
                se        = res$est_se
            )
        }) |> bind_rows()
    }) |> bind_rows()
}) |> bind_rows() -> results
```

- before doing this, test with `REPS = 5` and possibly fewer DGPs and estimators.
- use `future_map` instead of `map` and set up a parallel plan with `plan()`

## Analysis {.smaller}

```{r}
true_ATEs <- DGPS |> map_dbl(ATE)

results |>
    group_by(DGP, estimator) |>
    summarize(
        bias        = mean(est - true_ATEs[DGP]),
        true_se     = sd(est),
        mean_est_se = mean(se),
        rmse        = sqrt(mean((est - true_ATEs[DGP])^2)),
        coverage    = mean(
            (est - 1.96 * se <= true_ATEs[DGP]) &
            (true_ATEs[DGP] <= est + 1.96 * se)
        )
    )
```

From here you can use `tidyr`, `kable`, and `ggplot` to make tables and figures. 

## Tooling Demo

- **Positron** (files + notebook)
- **GitHub Copilot** in Positron Assistant

## What About Documentation & Testing?

- Don't bother, you'll have to rewrite it all tomorrow
- Simple comments here and there suffice
- When you write a new function, unit test it in a Jupyter notebook a few times informally
- GitHub? Sure. Just commit to main, don't worry about branches, PRs, etc.

# Iteration {background-color="#bababaff"}

## Problem!

> In your initial TMLE simulation, you find that the ML plug-in estimator outperforms TMLE and AIPW in terms of MSE. Debiasing improves bias, but at cost of variance.

* Why is this bad for your claim?
* What do you do?

## Solution

::: {.incremental}
* having a clear goal made it easy to know if result was "good" or "bad"
* understand theory: why is debiasing necessary?
* modify claim:
:::

. . .

> Like AIPW, TMLE improves the MSE of plugins in moderate sample sizes *when overlap is reasonable and initial estimates are biased* and allows for good coverage of Wald intervals based on IF SEs.

::: {.incremental}
* good claims are not overstated!
:::


# Conclusions

## How to do Simulations

```{dot}
//| echo: false
digraph {
  rankdir=TB
  node [shape=box, style="rounded,filled", fontname="Helvetica", fontsize=18, margin="0.3,0.2"]
  
  A [label="1. Paper Claims 🎯", fillcolor="#f88674ff"]
  B [label="2. Simulation Goals 🎯", fillcolor="#f88674ff"]
  C [label="3. Mockup Outline, Tables & Figures 📊", fillcolor="#f6cf71"]
  D [label="4. DGPs & Settings ⚙️", fillcolor="#66c5cc"]
  E [label="5. Implement ⚙️", fillcolor="#66c5cc"]
  
  A -> B -> C -> D -> E [penwidth=2, color=grey40]
  E -> A [penwidth=1.5, color=grey60, style=dashed, constraint=false]
}
```

## Top Tips

:::: {.columns}

::: {.column width="30%" style="background: #f88674ff"}
#### 🎯 Thoughtful Goals = Direction!

- practical utility
- key insight
- audience
:::

::: {.column width="30%" style="background: #f6cf71"}
#### 📊 Communication = Focus!

- structure + organize
- minimize cognitive load
:::

::: {.column width="30%" style="background: #66c5cc"}
#### ⚙️ Clean, Fast Sims = Productivity!

- simplest DGP
- minimize compute
- modular code
:::

::::